
---
title: "Semi-Competing Risk profiles - Restricted Mean Time Lost"
output: html_document
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

# {.tabset .tabset-fade .tabset-pills} 

## Setup

```{r, echo=T,message=F,warning=F,results='hide'}
library(tidyverse)
library(knitr)
library(ggpubr)

library(survival)

library(SemiCompRisks)
library(SemiCompRisksFreq)
library(splines2)


# use Harrison's color coding 
#color blind friendly colors from here: https://davidmathlogic.com/colorblind/#%23648FFF-%23785EF0-%23DC267F-%23FE6100-%23FFB000
cb_blue <- "#648FFF"; cb_red <- "#DC267F"; cb_purple <- "#785EF0"; cb_orange <- "#FE6100"; cb_grey <- "#CACACA"

#pink and green color-blind friendly color palette, with two colors (pink and green)
#each in two shades: light and dark
four_color_paired <- RColorBrewer::brewer.pal(n=4,name="PiYG")[c(3,4,2,1)]


#other colors used in plots
two_color_cb <- c(cb_blue,cb_red)

three_color <- c("dodgerblue","firebrick3","purple3")
three_color_cb <- c(cb_blue,cb_red,cb_purple)

#take three cb colors and reduce V from 100 to 60
three_color_cb_dark <- c("#3b5699", "#751342", "#453589")


four_color <- c("lightgray","firebrick3","purple3","dodgerblue")
four_color_cb <- c(cb_grey,cb_red,cb_purple,cb_blue)
four_color_forest <- c("dodgerblue","firebrick3","purple3","magenta")
four_color_forest_cb <- c(cb_blue,cb_red,cb_purple,cb_orange)
five_color <- c("lightgray","firebrick3","magenta","purple3","dodgerblue")
five_color_cb <- c(cb_grey,cb_red,cb_orange,cb_purple,cb_blue)

four_color_grey <- c("grey80","grey60","grey40","grey20")
five_color_grey <- c("grey80","grey60","grey40","grey20","black")

# RColorBrewer::display.brewer.all(n=4,colorblindFriendly = TRUE)
# color-blind friendly categorical colors
three_color_qual <- RColorBrewer::brewer.pal(n=3,name="Set2")
four_color_qual <- RColorBrewer::brewer.pal(n=4,name="Dark2")
five_color_qual <- RColorBrewer::brewer.pal(n=5,name="Dark2")

```

## Simulation of semicompeting risk data 
We simulate a training data set (n = 2000) and a validation data set (n = 1000). 
```{r, echo=T,message=F,warning=F,results='hide'}
set.seed(1234)

# sample size for simulation data set 
n <- 3000

# 3 time-independent baseline Gaussian covariates 
x_sim <- matrix(rnorm(3 * n), ncol = 3)
colnames(x_sim) <- c("x1","x2","x3")


# h3tv_knots define the intervals for each different effect of t1 on h3
h3tv_knots_temp <- c(3, 8, 15.5)

# true parameter values 
# vectors of true regression 
# beta1_true <- c(0.25,0.6,-0.3)
# beta2_true <- c(0.4,0.75,-0.4)
# beta3_true <- c(0.7,0.9,-0.9)

beta1_true <- c(0.1, 0.1, -0.2)

beta2_true <- c(0.4,0.75,-0.4)
beta3_true <- c(0.7,0.9,-0.9)

# true baseline hazard 
alpha1_true <- 1
alpha2_true <- 1
alpha3_true <- 1


kappa1_true <- exp(-3.0)
kappa2_true <- exp(-3.9)
kappa3_true <- exp(-3.3)

# mean for weibull 
wb_moments <- function(alpha, kappa) {
   shape <- alpha
   scale <- exp(-(log(kappa)))
   
   mean <- scale * gamma(1 + 1 / alpha)
   var <- scale^2 * (gamma(1 + 2 / alpha) - gamma(1 + 1 / alpha)^2)
   
   return(list(mean = mean, 
               sd = var %>% sqrt()))
}

wb_moments(alpha = alpha1_true, 
           kappa = kappa1_true)
wb_moments(alpha = alpha2_true, 
           kappa = kappa2_true)
wb_moments(alpha = alpha3_true, 
           kappa = kappa3_true)


theta_true <- 1
h3tv_degree <- 0 
h3tv_knots <- c(0,h3tv_knots_temp,Inf)
beta3tv_true <- c(0, 0, 0)

# no clustering 
simDat <- SemiCompRisksFreq::simID2(x1 = x_sim, 
                                    x2 = x_sim, 
                                    x3 = x_sim, 
                                    
                                    # vectors of true regression 
                                    beta1.true = beta1_true,
                                    beta2.true = beta2_true,
                                    beta3.true = beta3_true, 
                                    
                                    # true baseline hazard 
                                    alpha1.true = alpha1_true, 
                                    alpha2.true = alpha2_true, 
                                    alpha3.true = alpha3_true, 
                                    
                                    kappa1.true = kappa1_true,
                                    kappa2.true = kappa2_true,
                                    kappa3.true = kappa3_true,
                                    
                                    # Weibull baseline hazard
                                    hazard = "wb",
                                    # Semi-Markov model 
                                    model = "sm", 
                                    # Shared gamma frailty 
                                    frailty_type = "gamma", 
                                    theta.true = theta_true,
                                    
                                    # effect of time-to non-terminal event on terminal event 
                                    h3tv_degree = h3tv_degree, 
                                    h3tv_knots = h3tv_knots,
                                    beta3tv.true = beta3tv_true, 
                                    
                                    # censoring / upper max
                                    cens = c(100, 100)
                                    # cens = c(0, 0)
                                    )


# list of true parameters
paramList <- list(x1 = x_sim, 
                  x2 = x_sim, 
                  x3 = x_sim, 
                  
                  # vectors of true regression 
                  beta1.true = beta1_true,
                  beta2.true = beta2_true,
                  beta3.true = beta3_true, 
                  
                  # true baseline hazard 
                  alpha1.true = alpha1_true, 
                  alpha2.true = alpha2_true, 
                  alpha3.true = alpha3_true, 
                  
                  kappa1.true = kappa1_true,
                  kappa2.true = kappa2_true,
                  kappa3.true = kappa3_true,
                  
                  theta = theta_true, 
                  
                  h3tv_degree = h3tv_degree, 
                  h3tv_knots = h3tv_knots,
                  beta3tv.true = beta3tv_true)


simDat %>% head()
simDat %>% colnames()
simDat$y1 %>% summary()
simDat$y2 %>% summary()

table(simDat$delta1, simDat$delta2)

# add the x data to the data frame
simDat <- simDat %>%  
  cbind(id = 1:nrow(simDat), 
        x_sim)

simDat %>% head()

# define basis function for piecewise effect of non-terminal time on terminal time 
h3_b_fun <- function(x) splines2::bSpline(x,
                                          knots = h3tv_knots_temp,
                                          Boundary.knots = c(0,Inf),
                                          degree = 0,intercept = FALSE)

# define time and event variables to be used below
simDat$time1 <- simDat$y1
simDat$time2 <- simDat$y2
simDat$event1 <- simDat$delta1
simDat$event2 <- simDat$delta2
# define "sojourn time" explicitly
simDat$sojourn <- simDat$time2 - simDat$time1

# create event flags 
# define version of terminal flag that treats non-terminal event as censoring
scrData$event2_cr <- ifelse(scrData$event1 == 1, 0, 
                            scrData$event2)

# define version of event flag that is 0 for censoring, 1 for non-terminal, 2 for terminal
simDat$event_cr_num <- ifelse(simDat$event1 == 1, 1, 
                              ifelse(simDat$event2 == 1, 2, 0))
# reformat to factor variable 
simDat$event_cr_fct <- factor(simDat$event_cr_num,levels = c("0","1","2"),
                               labels = c("cens","nonterm","term"))

# make a variable with 4 values corresponding to "observed outcome" categories
outcome_vec <- numeric(nrow(simDat))
outcome_vec[scrData$event1==1 & scrData$event2==0] <- 1
outcome_vec[scrData$event1==0 & scrData$event2==1] <- 2
outcome_vec[scrData$event1==1 & scrData$event2==1] <- 3
simDat$outcome_cat <- factor(as.character(outcome_vec), levels=c("0","1","2","3"),
                              labels=c("neither","nonterm_only","term_only","both"))

# create a single factor variable corresponding to the cut points of t1 effect on h3
simDat$t1_cat <- cut(simDat$time1, 
                     breaks=c(0,h3tv_knots_temp,Inf),
                     right = FALSE, include.lowest = TRUE)

# render x1 a binary covariate 
simDat$x1_bin <- as.numeric(simDat$x1>0)

# separate training and validation data set
random_id <- sample(seq(1, nrow(simDat)), 
                    size = 2000, 
                    replace = FALSE)

trainDat <- simDat[random_id, ]
valDat <- simDat[-random_id, ]


#lastly, make a bonus dataset that "record-splits" the data into separate
#records for time following non-terminal event, with two different "event variables"
#one that just reports the occurrence of non-terminal and terminal events
#one that separates "terminal without non-terminal" and "terminal following non-terminal"

#rows corresponding to "first" transition
# temp_dat1 <- simDat
# temp_dat1$start_days <- 0
# temp_dat1$stop_days <- simDat$time1
# temp_dat1$ad_death_msm1 <- simDat$event_cr_num
# temp_dat1$ad_death_msm2 <- simDat$event_cr_num
# #rows corresponding to "second" transition if applicable
# temp_dat2 <- simDat[simDat$event1==1,]
# temp_dat2$start_days <- temp_dat2$time1
# temp_dat2$stop_days <- temp_dat2$time2
# temp_dat2$ad_death_msm1 <- ifelse(temp_dat2$event2,2,0)
# temp_dat2$ad_death_msm2 <- ifelse(temp_dat2$event2,3,0)
# #combine and sort data
# simDat_long_temp <- rbind(temp_dat1,temp_dat2)
# simDat_long <- simDat_long_temp[order(simDat_long_temp$id,simDat_long_temp$start_days),]
# simDat_long$ad_death_msm1 <- factor(simDat_long$ad_death_msm1,levels=c("0","1","2"),labels=c("cens","nonterm","term"))
# simDat_long$ad_death_msm2 <- factor(simDat_long$ad_death_msm2,levels=c("0","1","2","3"),labels=c("cens","nonterm_only","term_wo_nonterm","term_w_nonterm"))

```


### Visualize scatter plot for simulated data

```{r, echo=T,message=F,warning=F,results='show'}
#Empirical "Density" Plot
scat_plot <- ggplot(data = trainDat %>% filter(event1 == 1),
                    mapping = aes(x=time1,
                                  y=time2, 
                                  color=as.factor(event2))) +
  labs(x = "Time to non-terminal event", 
       y = "Time to terminal event", 
       color = "Terminal event happened?") + 
  xlim(0, 102) +
  ylim(0, 102) +
  geom_point(alpha=0.25, size=1) + 
  theme_bw() +
  geom_abline(slope=1) + 
  theme(legend.position = "bottom")

#histogram of terminal only times
hist_plot <- ggplot(data=trainDat %>% filter(event1 == 0),
                    mapping = aes(x=time2,
                                  fill=as.factor(event2))) + 
  xlim(0,102) +
  geom_histogram(binwidth = 1,
                 col="white",
                 linewidth=0.25) +
  theme_bw() + 
  coord_flip() + 
  labs(x = "Time to terminal event without non-terminal event", 
       y = "Count", 
       fill = "Terminal event happened?") +
  theme(legend.position = "bottom", 
        legend.title = element_blank())
#combined "empirical joint density"
cowplot::plot_grid(scat_plot,hist_plot, 
                   nrow = 1, rel_widths = c(7,3))

```

### Event rate for terminal and non-terminal event  
At the end of the observation period, without any intermediate censoring, only administrative censoring at 100, the event rate is 

```{r}
table(simDat$event1, simDat$event2) %>% 
  addmargins()

table(simDat$event1, simDat$event2) %>%
  prop.table() %>% 
  addmargins() %>% 
  round(3)

```

## Save data
```{r}
save(trainDat, file = "Data/SimData_Training.RData")
save(valDat, file = "Data/SimData_Validation.RData")

save(paramList, file = "Data/TrueParameter_List.RData")
```

