
---
title: "Semi-Competing Risk profiles - Restricted Mean Time Lost for Weibull baseline hazard"
output: html_document
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

# {.tabset .tabset-fade .tabset-pills} 

## Setup

```{r, echo=T,message=F,warning=F,results='hide'}
library(tidyverse)
library(knitr)
library(ggpubr)

library(survival)

library(SemiCompRisks)
library(SemiCompRisksFreq)
library(splines2)


# use Harrison's color coding 
#color blind friendly colors from here: https://davidmathlogic.com/colorblind/#%23648FFF-%23785EF0-%23DC267F-%23FE6100-%23FFB000
cb_blue <- "#648FFF"; cb_red <- "#DC267F"; cb_purple <- "#785EF0"; cb_orange <- "#FE6100"; cb_grey <- "#CACACA"

#pink and green color-blind friendly color palette, with two colors (pink and green)
#each in two shades: light and dark
four_color_paired <- RColorBrewer::brewer.pal(n=4,name="PiYG")[c(3,4,2,1)]


#other colors used in plots
two_color_cb <- c(cb_blue,cb_red)

three_color <- c("dodgerblue","firebrick3","purple3")
three_color_cb <- c(cb_blue,cb_red,cb_purple)

#take three cb colors and reduce V from 100 to 60
three_color_cb_dark <- c("#3b5699", "#751342", "#453589")


four_color <- c("lightgray","firebrick3","purple3","dodgerblue")
four_color_cb <- c(cb_grey,cb_red,cb_purple,cb_blue)
four_color_forest <- c("dodgerblue","firebrick3","purple3","magenta")
four_color_forest_cb <- c(cb_blue,cb_red,cb_purple,cb_orange)
five_color <- c("lightgray","firebrick3","magenta","purple3","dodgerblue")
five_color_cb <- c(cb_grey,cb_red,cb_orange,cb_purple,cb_blue)

four_color_grey <- c("grey80","grey60","grey40","grey20")
five_color_grey <- c("grey80","grey60","grey40","grey20","black")

# RColorBrewer::display.brewer.all(n=4,colorblindFriendly = TRUE)
# color-blind friendly categorical colors
three_color_qual <- RColorBrewer::brewer.pal(n=3,name="Set2")
four_color_qual <- RColorBrewer::brewer.pal(n=4,name="Dark2")
five_color_qual <- RColorBrewer::brewer.pal(n=5,name="Dark2")

```

## Functions 
```{r, echo=T,message=F,warning=F,results='hide'}
h3tv_knots_temp <- c(3, 8, 15.5)
# define basis function for piecewise effect of non-terminal time on terminal time 
h3_b_fun <- function(x) splines2::bSpline(x,
                                          knots = h3tv_knots_temp,
                                          Boundary.knots = c(0,Inf),
                                          degree = 0,intercept = FALSE)
# function to plot risk profile 
plot_risk_profile <- function(tseq_pred,
                              ValPred, 
                              i, 
                              colors = four_color_cb) {
  length_tseq <- length(tseq_pred)
  
  # format data frame 
  plot_frame <- data.frame(
    Time = rep(tseq_pred, 4),
    Probability = c(ValPred$p_neither[, i], 
                    ValPred$p_term_only[, i],
                    ValPred$p_both[, i],
                    ValPred$p_nonterm_only[, i]),
    Outcome=factor(x = c(rep("Neither",length_tseq),
                         rep("Terminal Only",length_tseq),
                         rep("Both",length_tseq),
                         rep("Nonterminal Only",length_tseq)),
                   levels = c("Neither",
                              "Terminal Only",
                              "Both",
                              "Nonterminal Only"))
  )
  
    g <- ggplot(plot_frame, 
              aes(x = Time, 
                  y = Probability)) +
    geom_area(aes(colour = Outcome, 
                  fill = Outcome)) +
    scale_color_manual(values = colors) +
    scale_fill_manual(values = colors) + 
    theme_bw() +
    labs(title = paste0("Patient ", i, collapse = ""))
    
    if(valDat[i, ]$event1 == 1) {
    trueT1 <- valDat[i, ]$time1
    
    g <- g +
      geom_vline(aes(xintercept = trueT1), 
                 linetype = "dashed", 
                 color = "red")
    
  }
  if(valDat[i, ]$event2 == 1) {
    trueT2 <- valDat[i, ]$time2
    
     g <- g +
      geom_vline(aes(xintercept = trueT2), 
                 linetype = "dashed", 
                 color = "black")
  }

    return(g)
}
```

## Load Data 
```{r, echo=T,message=F,warning=F,results='show'}
# load data file 
load(file = "Data/SimData_Training.RData")
load(file = "Data/SimData_Validation.RData")


load(file = "Data/TrueParameter_List.RData")

# extract true parameters
x_sim <- paramList$x1

beta1_true <- paramList$beta1.true
beta2_true <- paramList$beta2.true
beta3_true <- paramList$beta3.true

alpha1_true <- paramList$alpha1.true
alpha2_true <- paramList$alpha2.true
alpha3_true <- paramList$alpha3.true


kappa1_true <- paramList$kappa1.true
kappa2_true <- paramList$kappa2.true
kappa3_true <- paramList$kappa3.true

h3tv_degree <- paramList$h3tv_degree 
h3tv_knots <- paramList$h3tv_knots
beta3tv_true <- paramList$beta3tv.true


theta_true <- paramList$theta

```


## Shared frailty Illness-Death model with Weibull baseline hazard 

```{r, echo=T,message=F,warning=F,results='show'}
# define formula 
form_temp_no_t1cat <- Formula::Formula(time1 + event1 | time2 + event2 ~ x1 + x2 + x3 | x1 + x2 + x3 | x1 + x2 + x3 )

SFID_no_t1cat <- SemiCompRisksFreq::FreqID_HReg2(Formula = form_temp_no_t1cat, 
                                                 data = trainDat,
                                                 hazard = "wb", 
                                                 model = "semi-Markov",
                                                 # number of baseline parameters 
                                                 nP0 = c(2,2,2),
                                                 frailty = TRUE, 
                                                 optim_method = "BFGS",
                                                 extra_starts = 0)

SFID_no_t1cat %>% summary()

```

### Visualize fitted shared frailty Illness-Death model 
We are restricting ourselves to a time window 0 to 100. 
Since no censoring was modeled, the terminal event time for some patients might be very high. 

```{r, echo=T,message=F,warning=F,results='show'}
# estimate for a certain time window 
tseq_pred <- seq(0, 100)

SFID_no_t1cat_pred <- predict(SFID_no_t1cat, 
                              tseq = tseq_pred)

plot(SFID_no_t1cat_pred, plot.est = "Haz")
plot(SFID_no_t1cat_pred, plot.est = "CumHaz")
plot(SFID_no_t1cat_pred, plot.est = "Surv")

```


### Forest plot for model coefficients 
The covariate coefficients for the marginal and conditional hazard functions are estimated with high accuracy. 

```{r, eval=T,echo=T}
SFID_coef <- summary(SFID_no_t1cat)$coef_long %>% 
  as.data.frame() %>% 
  mutate(name = c("beta1.1", "beta1.2", "beta1.3", 
                  "beta2.1", "beta2.2", "beta2.3", 
                  "beta3.1", "beta3.2", "beta3.3"), 
         haz = c(rep("non-terminal", 3), 
                 rep("terminal", 3), 
                 rep("terminal | non-terminal", 3)))

SFID_coef_true <- data.frame(name = c("beta1.1", "beta1.2", "beta1.3", 
                                      "beta2.1", "beta2.2", "beta2.3", 
                                      "beta3.1", "beta3.2", "beta3.3"), 
                             true_coef = c(beta1_true, beta2_true, beta3_true), 
                             haz = c(rep("non-terminal", 3), 
                                     rep("terminal", 3), 
                                     rep("terminal | non-terminal", 3)))

ggplot(data = SFID_coef, 
       aes(x = name, 
           y = beta, 
           color = haz %>% as.factor())) +
  geom_point() +
  geom_point(data = SFID_coef_true, 
             aes(x = name, 
                 y = true_coef, 
                 color = haz %>% as.factor(), 
                 shape = "True coef."),
             size = 4) +
  geom_errorbar(aes(ymin = beta - 1.96 * SE, 
                    ymax = beta + 1.96 * SE)) +
  coord_flip() +
  theme_bw() +
  theme(legend.position = "bottom") +
  labs(y = "Coefficients", 
       x = "", 
       color = "", 
       shape = "") +
  scale_shape_manual(values = c("True coef." = 8))
```
### Forest plot for Baseline hazard coefficients


```{r}
# theta frailty parameter 
# kappa / alpha WB baseline hazard parameter

bs_haz_coef <- summary(SFID_no_t1cat)$h0_long %>% 
  exp() %>% # reported on a log scale; monotone increasing so transformation of CI possible; alternatively delta method 
  as.data.frame() %>% 
  mutate(name = c("kappa1", "alpha1", 
                  "kappa2", "alpha2", 
                  "kappa12", "alpha12"), 
         haz = c(rep("non-terminal", 2), 
                 rep("terminal", 2), 
                 rep("terminal | non-terminal", 2)))

theta_coef <- summary(SFID_no_t1cat)$theta %>% t() %>%  
  as.data.frame() %>% 
  dplyr::select(theta, SE, LL, UL) %>% 
  mutate(name = "theta", 
         haz = "frailty") %>% 
    rename(beta = theta)

bs_haz_coef <- rbind(bs_haz_coef, 
                     theta_coef)

bs_haz_coef_true <- data.frame(name = c("kappa1", "alpha1", 
                                        "kappa2", "alpha2", 
                                        "kappa12", "alpha12", 
                                        "theta"), 
                               beta = c(kappa1_true, alpha1_true, 
                                        kappa2_true, alpha2_true, 
                                        kappa3_true, alpha3_true, 
                                        theta_true), 
         haz = c(rep("non-terminal", 2), 
                 rep("terminal", 2), 
                 rep("terminal | non-terminal", 2), 
                 "frailty"))


ggplot(data = bs_haz_coef, 
       aes(x = name, 
           y = beta, 
           color = haz %>% as.factor())) +
  geom_point() +
  geom_point(data = bs_haz_coef_true, 
             aes(x = name, 
                 y = beta, 
                 color = haz %>% as.factor(), 
                 shape = "True coef."),
             size = 4) +
  geom_errorbar(aes(ymin = beta - 1.96 * SE, 
                    ymax = beta + 1.96 * SE)) +
  coord_flip() +
  theme_bw() +
  theme(legend.position = "bottom") +
  labs(y = "Coefficients", 
       x = "", 
       color = "", 
       shape = "") +
  scale_shape_manual(values = c("True coef." = 8))
```


### Baseline hazard function 
The baseline hazard is fit assuming a Weibull form, which aligns with the true form of the baseline hazard function underlying the simulation. 

We need to be careful regarding the different parametrizations of the Weibull distribution in the stats package and the SemiCompRiskFreq package. 


The standard R parametrization is
\begin{align}
f(x) & = \frac{\alpha}{\sigma} \left( \frac{x}{\sigma} \right)^{\alpha - 1} \exp\left( - \frac{x}{\sigma}^{\alpha} \right)
\end{align}

The parametrization in the SemiCompRiskFreq package is 
\begin{align}
f(x) & = \kappa \alpha x^{\alpha-1} \exp(-\kappa x^{\alpha})
\end{align}
The estimates are reported on a log scale. 
\begin{align}
\alpha & = \exp(lalpha) \\
\sigma & = \exp(-lkappa / \alpha)
\end{align}

```{r, eval=T,echo=T}
bs_hazard_1 <- data.frame(param = rep(paste0("alpha1 = ", alpha1_true, 
                                            ", kappa1 = ", kappa1_true %>% round(2)), 
                                      length(tseq_pred)), 
                          type = "h1", 
                          time = tseq_pred, 
                          bs_hazard = dweibull(x = tseq_pred, 
                                               shape = alpha1_true, 
                                               scale = exp(-log(kappa1_true) / alpha1_true)))


# baseline hazard for terminal event 
bs_hazard_2 <- data.frame(param = rep(paste0("alpha2 = ", alpha2_true, 
                                            ", kappa2 = ", kappa2_true %>% round(2)), 
                                      length(tseq_pred)), 
                          type = "h2", 
                          time = tseq_pred, 
                          bs_hazard = dweibull(x = tseq_pred, 
                                               shape = alpha2_true, 
                                               scale = exp(-log(kappa2_true) / alpha2_true)))

# conditional baseline hazard 
bs_hazard_3 <- data.frame(param = rep(paste0("alpha3 = ", alpha3_true, 
                                            ", kappa3 = ", kappa3_true %>% round(2)), 
                                      length(tseq_pred)), 
                          type = "h3", 
                          time = tseq_pred, 
                          bs_hazard = dweibull(x = tseq_pred, 
                                               shape = alpha3_true, 
                                               scale = exp(-log(kappa3_true) / alpha3_true)))

true_bs_hazards <- rbind(bs_hazard_1, 
                         bs_hazard_2,
                         bs_hazard_3)

# estimated baseline hazard function 
bs_est_hazard_1 <- data.frame(type = "h1", 
                          time = tseq_pred, 
                          bs_hazard = dweibull(x = tseq_pred, 
                                               shape = (bs_haz_coef %>%
                                                 filter(name == "alpha1") %>%
                                                 pull(beta)), 
                                               scale = exp(-log(bs_haz_coef %>%
                                                 filter(name == "kappa1") %>%
                                                 pull(beta)) / (bs_haz_coef %>%
                                                 filter(name == "alpha1") %>%
                                                 pull(beta))))
                                               )


# baseline hazard for terminal event 
bs_est_hazard_2 <- data.frame(type = "h2", 
                          time = tseq_pred, 
                          bs_hazard = dweibull(x = tseq_pred, 
                                               shape = (bs_haz_coef %>%
                                                 filter(name == "alpha2") %>%
                                                 pull(beta)), 
                                               scale = exp(-log(bs_haz_coef %>%
                                                 filter(name == "kappa2") %>%
                                                 pull(beta)) / (bs_haz_coef %>%
                                                 filter(name == "alpha2") %>%
                                                 pull(beta))))
                                               )

# conditional baseline hazard 
bs_est_hazard_3 <- data.frame(type = "h3", 
                          time = tseq_pred, 
                          bs_hazard = dweibull(x = tseq_pred, 
                                               shape = (bs_haz_coef %>%
                                                 filter(name == "alpha12") %>%
                                                 pull(beta)), 
                                               scale = exp(-log(bs_haz_coef %>%
                                                 filter(name == "kappa12") %>%
                                                 pull(beta)) / (bs_haz_coef %>%
                                                 filter(name == "alpha12") %>%
                                                 pull(beta))))
                                               )


est_bs_hazards <- rbind(bs_est_hazard_1, 
                        bs_est_hazard_2, 
                        bs_est_hazard_3)

ggplot(data = true_bs_hazards, 
       aes(x = time,
           y = bs_hazard, 
           color = type)) +
  geom_line(linetype = "dashed") +
  geom_line(data = est_bs_hazards,
            aes(x = time,
                y = bs_hazard,
                color = type)) +
  theme_bw() +
  theme(legend.position = "bottom") +
  labs(y = "baseline hazard function", 
       color = "")

```



