---
title: "Simulation of discrete time data"
output: html_document
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

# {.tabset .tabset-fade .tabset-pills} 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Library -----------------------------------------------------------------
# Functions
source("Data_generating_mechanism.R")
source("Simulation_analysis_helpers.R")

library(tidyverse)

# Set seed
set.seed(123)

```

## Simulation 0
No association between longitudinal marker and mortality. 
Longitudinal marker has linear trend and independent error over time, no autoregressive component. 
Baseline mortality risk is time- and covariate-independent, i.e. constant over time and patients. 
Mortality is independent of the longitudinal trajectory. 


```{r}
# Simulation 1 --------------------------------------------------------------
# no association between marker and mortality
times <- seq(0, 14, 0.5)
n <- 100000


## MORTALITY
lambda.t <- rep(logit(0.01), length(times) - 1)
# positive effect of marker on death probability
theta.t <- 0
varphi.t <- runif(n = length(times) - 1, min = 0, max = 0)
xi.t <- c(0, 0, 0)

## LONGITUDINAL MARKER
zeta.long <- 0.5*(times)
eta.long <- rep(0, length(times) - 1)
beta.long <- c(-0.2, 0.3, -0.4)

```

### Baseline mortality risk 
```{r}
# plot baseline mortality risk
plot_baseline_mortality_risk(times,
                             lambda.t)

```

### Longitudinal trajectory trend
```{r}
# plot longitudinal mean
plot_longitudinal_trend(times,
                        zeta.long)
```

### Patients dying? 
```{r}
# Simulation of outcome data
sim_dat <- SimulateJointLongData(n.sample = n,
                                 times = times,
                                 p = 3,
                                 lambda.t = lambda.t,
                                 theta.t = theta.t,
                                 varphi.t = varphi.t,
                                 xi.t = xi.t,
                                 zeta.long = zeta.long,
                                 eta.long = eta.long,
                                 beta.long = beta.long,

                                 sd.long.trajectory = 1)

sim_data <- sim_dat$df.return %>%
  ungroup()

# How many patients die over the entire study duration?
sim_data %>%
  group_by(ID) %>%
  reframe(dies = any(Death == 1)) %>%
  pull(dies) %>%
  table()

# time of death
sim_data %>%
  group_by(ID) %>%
  dplyr::filter(Death == 1) %>%
  reframe(death_time = min(Time)) %>%
  pull(death_time) %>%
  table()

```

### Patients at risk 

```{r}
# patients at risk over time
plot_patients_at_risk(sim_data)
```

### Expected longitudinal marker trajectory for mortal and immortal cohort 

```{r}
# expected longitudinal trajectory over time
plot_immortal_mortal_longitudinal_trajectory(sim_data)
```

### Heterogeneity in longitudinal marker trajectory 

```{r}
# patient-specific longitudinal trajectories over time
plot_heterogeneity(data = sim_data)
```




## Simulation 1
No association between longitudinal marker and mortality. 
Longitudinal marker has linear trend and independent error over time, no autoregressive component. 
Baseline mortality risk is time-independent but does depend on covariates. 
Mortality is independent of the longitudinal trajectory. 


```{r}
# Simulation 1 --------------------------------------------------------------
# no association between marker and mortality
times <- seq(0, 14, 0.5)
n <- 100000


## MORTALITY
lambda.t <- rep(logit(0.01), length(times) - 1)
# positive effect of marker on death probability
theta.t <- 0
varphi.t <- rep(0, length(times) - 1)
xi.t <- c(1, 2, 3)

## LONGITUDINAL MARKER
zeta.long <- 0.5*(times)
eta.long <- rep(0, length(times) - 1)
beta.long <- c(1, 2, 3)

```

### Baseline mortality risk 
```{r}
# plot baseline mortality risk
plot_baseline_mortality_risk(times,
                             lambda.t)

```

### Longitudinal trajectory trend
```{r}
# plot longitudinal mean
plot_longitudinal_trend(times,
                        zeta.long)
```

### Patients dying? 
```{r}
# Simulation of outcome data
sim_dat <- SimulateJointLongData(n.sample = n,
                                 times = times,
                                 p = 3,
                                 lambda.t = lambda.t,
                                 theta.t = theta.t,
                                 varphi.t = varphi.t,
                                 xi.t = xi.t,
                                 zeta.long = zeta.long,
                                 eta.long = eta.long,
                                 beta.long = beta.long,

                                 sd.long.trajectory = 0.1)

sim_data <- sim_dat$df.return %>%
  ungroup()

# How many patients die over the entire study duration?
sim_data %>%
  group_by(ID) %>%
  reframe(dies = any(Death == 1)) %>%
  pull(dies) %>%
  table()

# time of death
sim_data %>%
  group_by(ID) %>%
  dplyr::filter(Death == 1) %>%
  reframe(death_time = min(Time)) %>%
  pull(death_time) %>%
  table()

```

### Patients at risk 

```{r}
# patients at risk over time
plot_patients_at_risk(sim_data)
```

### Expected longitudinal marker trajectory for mortal and immortal cohort 

```{r}
# expected longitudinal trajectory over time
plot_immortal_mortal_longitudinal_trajectory(sim_data)
```

### Heterogeneity in longitudinal marker trajectory 

```{r}
# patient-specific longitudinal trajectories over time
plot_heterogeneity(data = sim_data)
```

## Simulation 2
The longitudinal marker trajectory is linear and has no autoregressive component. 
It is covariate dependent. 
The marker and mortality are positively associated based on a global dependence parameter, no local dependence changes over time. 
Mortality does not depend on covariates, i.e. the marker is the only component introducing association. 

```{r}
# Simulation 2 ------------------------------------------------------------
# linear trend in longitudinal marker trajectory
# not autoregressive component
# constant baseline mortality risk

times <- seq(0, 14, 0.5)
n <- 100000


## MORTALITY
lambda.t <- rep(logit(0.01), length(times) - 1)
# positive effect of marker on death probability
theta.t <- 6
varphi.t <- rep(0, length(times) - 1)
xi.t <- c(0, 0, 0)

## LONGITUDINAL MARKER
zeta.long <- 0.05*(times)
eta.long <- rep(0, length(times) - 1)
beta.long <- c(-0.2, 0.3, -0.4)

```

### Baseline mortality risk 
```{r}
# plot baseline mortality risk
plot_baseline_mortality_risk(times,
                             lambda.t)

```

### Longitudinal trajectory trend
```{r}
# plot longitudinal mean
plot_longitudinal_trend(times,
                        zeta.long)
```

### Patients dying? 
```{r}
# Simulation of outcome data
sim_dat <- SimulateJointLongData(n.sample = n,
                                 times = times,
                                 p = 3,
                                 lambda.t = lambda.t,
                                 theta.t = theta.t,
                                 varphi.t = varphi.t,
                                 xi.t = xi.t,
                                 zeta.long = zeta.long,
                                 eta.long = eta.long,
                                 beta.long = beta.long,

                                 sd.long.trajectory = 1)

sim_data <- sim_dat$df.return %>%
  ungroup()

# How many patients die over the entire study duration?
sim_data %>%
  group_by(ID) %>%
  reframe(dies = any(Death == 1)) %>%
  pull(dies) %>%
  table()

# time of death
sim_data %>%
  group_by(ID) %>%
  dplyr::filter(Death == 1) %>%
  reframe(death_time = min(Time)) %>%
  pull(death_time) %>%
  table()

```

### Patients at risk 

```{r}
# patients at risk over time
plot_patients_at_risk(sim_data)
```

### Expected longitudinal marker trajectory for mortal and immortal cohort 

```{r}
# expected longitudinal trajectory over time
plot_immortal_mortal_longitudinal_trajectory(sim_data)
```

### Heterogeneity in longitudinal marker trajectory 

```{r}
# patient-specific longitudinal trajectories over time
plot_heterogeneity(data = sim_data)
```


## Simulation 3.1
The mortality risk and longitudinal marker trajectory have a quadratic trend. 
They are positively associated globally, without local changes. 

The mortality risk is not covariate dependent. 

```{r}
# Simulation 3 ------------------------------------------------------------
# time sequence
times <- seq(0, 14, 0.5)
# size of sample
n <- 100000


## MORTALITY
lambda.t <- (logit(times*0.0002  + 0.0001*(times^2) + 0.005))[-1]
# positive effect of marker on death probability
theta.t <- 5
varphi.t <- rep(0, length(times) - 1)
xi.t <- c(0, 0, 0)

## LONGITUDINAL MARKER
zeta.long <- 0.005*(times)^2
eta.long <- rep(0, length(times) - 1)
beta.long <- c(-0.2, 0.3, -0.4)
```

### Baseline mortality risk 
```{r}
# plot baseline mortality risk
plot_baseline_mortality_risk(times,
                             lambda.t)

```

### Longitudinal trajectory trend
```{r}
# plot longitudinal mean
plot_longitudinal_trend(times,
                        zeta.long)
```

### Patients dying? 
```{r}
# Simulation of outcome data
sim_dat <- SimulateJointLongData(n.sample = n,
                                 times = times,
                                 p = 3,
                                 lambda.t = lambda.t,
                                 theta.t = theta.t,
                                 varphi.t = varphi.t,
                                 xi.t = xi.t,
                                 zeta.long = zeta.long,
                                 eta.long = eta.long,
                                 beta.long = beta.long,

                                 sd.long.trajectory = 0.1)

sim_data <- sim_dat$df.return %>%
  ungroup()

# How many patients die over the entire study duration?
sim_data %>%
  group_by(ID) %>%
  reframe(dies = any(Death == 1)) %>%
  pull(dies) %>%
  table()

# time of death
sim_data %>%
  group_by(ID) %>%
  dplyr::filter(Death == 1) %>%
  reframe(death_time = min(Time)) %>%
  pull(death_time) %>%
  table()

```

### Patients at risk 

```{r}
# patients at risk over time
plot_patients_at_risk(sim_data)
```

### Expected longitudinal marker trajectory for mortal and immortal cohort 

```{r}
# expected longitudinal trajectory over time
plot_immortal_mortal_longitudinal_trajectory(sim_data)
```

### Heterogeneity in longitudinal marker trajectory 

```{r}
# patient-specific longitudinal trajectories over time
plot_heterogeneity(data = sim_data)
```



## Simulation 3.2
The mortality risk and longitudinal marker trajectory have a quadratic trend. 
They are negatively associated globally, without local changes. 

The mortality risk is not covariate dependent. 

```{r}
# Simulation 3 ------------------------------------------------------------
# time sequence
times <- seq(0, 14, 0.5)
# size of sample
n <- 100000


## MORTALITY
lambda.t <- (logit(times*0.0002  + 0.0001*(times^2) + 0.005))[-1]
# positive effect of marker on death probability
theta.t <- -10
varphi.t <- rep(0, length(times) - 1)
xi.t <- c(0, 0, 0)

## LONGITUDINAL MARKER
zeta.long <- 0.005*(times)^2
eta.long <- rep(0, length(times) - 1)
beta.long <- c(-0.2, 0.3, -0.4)
```

### Baseline mortality risk 
```{r}
# plot baseline mortality risk
plot_baseline_mortality_risk(times,
                             lambda.t)

```

### Longitudinal trajectory trend
```{r}
# plot longitudinal mean
plot_longitudinal_trend(times,
                        zeta.long)
```

### Patients dying? 
```{r}
# Simulation of outcome data
sim_dat <- SimulateJointLongData(n.sample = n,
                                 times = times,
                                 p = 3,
                                 lambda.t = lambda.t,
                                 theta.t = theta.t,
                                 varphi.t = varphi.t,
                                 xi.t = xi.t,
                                 zeta.long = zeta.long,
                                 eta.long = eta.long,
                                 beta.long = beta.long,

                                 sd.long.trajectory = 0.1)

sim_data <- sim_dat$df.return %>%
  ungroup()

# How many patients die over the entire study duration?
sim_data %>%
  group_by(ID) %>%
  reframe(dies = any(Death == 1)) %>%
  pull(dies) %>%
  table()

# time of death
sim_data %>%
  group_by(ID) %>%
  dplyr::filter(Death == 1) %>%
  reframe(death_time = min(Time)) %>%
  pull(death_time) %>%
  table()

```

### Patients at risk 

```{r}
# patients at risk over time
plot_patients_at_risk(sim_data)
```

### Expected longitudinal marker trajectory for mortal and immortal cohort 

```{r}
# expected longitudinal trajectory over time
plot_immortal_mortal_longitudinal_trajectory(sim_data)
```

### Heterogeneity in longitudinal marker trajectory 

```{r}
# patient-specific longitudinal trajectories over time
plot_heterogeneity(data = sim_data)
```



## Simulation 4
The longitudinal marker and the mortality probability are positively associated and depend on the covariates. 
Both also have a quadratic trend in time.
```{r}
# Simulation 3 ------------------------------------------------------------
# time sequence
times <- seq(0, 14, 0.5)
# size of sample
n <- 100000


## MORTALITY
lambda.t <- (logit(times*0.0002  + 0.0001*(times^2) + 0.005))[-1]
# positive effect of marker on death probability
theta.t <- 5
varphi.t <- runif(n = length(times) - 1, min = 0, max = 0.5)
xi.t <- c(0.02, -0.04, -0.03)

## LONGITUDINAL MARKER
zeta.long <- 0.05*(times)^2
eta.long <- runif(n = length(times) - 1, min = 0.05, max = 0.09)
beta.long <- c(-0.2, 0.3, -0.4)
```

### Baseline mortality risk 
```{r}
# plot baseline mortality risk
plot_baseline_mortality_risk(times,
                             lambda.t)

```

### Longitudinal trajectory trend
```{r}
# plot longitudinal mean
plot_longitudinal_trend(times,
                        zeta.long)
```

### Patients dying? 
```{r}
# Simulation of outcome data
sim_dat <- SimulateJointLongData(n.sample = n,
                                 times = times,
                                 p = 3,
                                 lambda.t = lambda.t,
                                 theta.t = theta.t,
                                 varphi.t = varphi.t,
                                 xi.t = xi.t,
                                 zeta.long = zeta.long,
                                 eta.long = eta.long,
                                 beta.long = beta.long,

                                 sd.long.trajectory = 1)

sim_data <- sim_dat$df.return %>%
  ungroup()

# How many patients die over the entire study duration?
sim_data %>%
  group_by(ID) %>%
  reframe(dies = any(Death == 1)) %>%
  pull(dies) %>%
  table()

# time of death
sim_data %>%
  group_by(ID) %>%
  dplyr::filter(Death == 1) %>%
  reframe(death_time = min(Time)) %>%
  pull(death_time) %>%
  table()

```

### Patients at risk 

```{r}
# patients at risk over time
plot_patients_at_risk(sim_data)

```
### Variance across time for longitudinal marker
The conditional variance within each partition, conditional on the lag-1 longitudinal marker value and the baseline covariates, remain constant.
Considering the marginal variance across partitions, we expect a positive increase in variance as time progresses. 


```{r}
var_over_time <- sim_data %>% 
  group_by(Time) %>% 
  reframe(var = var(Marker))


ggplot(data = var_over_time, 
       aes(x = Time, 
           y = var)) +
  geom_line() +
  theme_bw()

estimate_ts_covariance(sim_data, 
                       k = 1)
estimate_ts_covariance(sim_data, 
                       k = 5)
estimate_ts_covariance(sim_data, 
                       k = 10)

```


### Expected longitudinal marker trajectory for mortal and immortal cohort 

```{r}
# expected longitudinal trajectory over time
plot_immortal_mortal_longitudinal_trajectory(sim_data)
```

### Heterogeneity in longitudinal marker trajectory 

```{r}
# patient-specific longitudinal trajectories over time
plot_heterogeneity(data = sim_data)
```



## Simulation 5.1
The longitudinal marker has no trend but a lag-1 autoregressive component. 
The baseline mortality is constant. 

```{r}
# Simulation 3 ------------------------------------------------------------
# time sequence
times <- seq(0, 30, 0.5)
# size of sample
n <- 100000


## MORTALITY
lambda.t <- rep(logit(0.01), length(times) - 1)
# positive effect of marker on death probability
theta.t <- 0.5
varphi.t <- runif(n = length(times) - 1, min = 0, max = 0.3)
xi.t <- c(0.02, -0.04, -0.03)

## LONGITUDINAL MARKER
zeta.long <- rep(0, length(times))
eta.long <- rep(0.5, length(times) - 1)
beta.long <- c(0, 0, 0)
```

### Baseline mortality risk 
```{r}
# plot baseline mortality risk
plot_baseline_mortality_risk(times,
                             lambda.t)

```

### Longitudinal trajectory trend
```{r}
# plot longitudinal mean
plot_longitudinal_trend(times,
                        zeta.long)
```

### Patients dying? 
```{r}
# Simulation of outcome data
sim_dat <- SimulateJointLongData(n.sample = n,
                                 times = times,
                                 p = 3,
                                 lambda.t = lambda.t,
                                 theta.t = theta.t,
                                 varphi.t = varphi.t,
                                 xi.t = xi.t,
                                 zeta.long = zeta.long,
                                 eta.long = eta.long,
                                 beta.long = beta.long,

                                 sd.long.trajectory = 1)

sim_data <- sim_dat$df.return %>%
  ungroup()

# How many patients die over the entire study duration?
sim_data %>%
  group_by(ID) %>%
  reframe(dies = any(Death == 1)) %>%
  pull(dies) %>%
  table()

# time of death
sim_data %>%
  group_by(ID) %>%
  dplyr::filter(Death == 1) %>%
  reframe(death_time = min(Time)) %>%
  pull(death_time) %>%
  table()

```

### Patients at risk 

```{r}
# patients at risk over time
plot_patients_at_risk(sim_data)

```

### Variance across time for longitudinal marker
The longitudinal trajectory is stationary, as visible from the constant covariance.  


```{r}
var_over_time <- sim_data %>% 
  group_by(Time) %>% 
  reframe(var = var(Marker))


ggplot(data = var_over_time, 
       aes(x = Time, 
           y = var)) +
  geom_line() +
  theme_bw()

# estimate_ts_covariance(sim_data, 
#                        k = 1)
# estimate_ts_covariance(sim_data, 
#                        k = 5)
# estimate_ts_covariance(sim_data, 
#                        k = 10)

```


### Expected longitudinal marker trajectory for mortal and immortal cohort 

```{r}
# expected longitudinal trajectory over time
plot_immortal_mortal_longitudinal_trajectory(sim_data)
```

### Heterogeneity in longitudinal marker trajectory 

```{r}
# patient-specific longitudinal trajectories over time
plot_heterogeneity(data = sim_data)
```



## Simulation 5.2
The longitudinal marker has no trend but a lag-1 autoregressive component. 
The baseline mortality is constant. 

```{r}
# Simulation 3 ------------------------------------------------------------
# time sequence
times <- seq(0, 30, 0.5)
# size of sample
n <- 100000


## MORTALITY
lambda.t <- rep(logit(0.01), length(times) - 1)
# positive effect of marker on death probability
theta.t <- 0.5
varphi.t <- runif(n = length(times) - 1, min = 0, max = 0.3)
xi.t <- c(0.02, -0.04, -0.03)

## LONGITUDINAL MARKER
zeta.long <- rep(0, length(times))
eta.long <- rep(1.2, length(times) - 1)
beta.long <- c(0, 0, 0)
```

### Baseline mortality risk 
```{r}
# plot baseline mortality risk
plot_baseline_mortality_risk(times,
                             lambda.t)

```

### Longitudinal trajectory trend
```{r}
# plot longitudinal mean
plot_longitudinal_trend(times,
                        zeta.long)
```

### Patients dying? 
```{r}
# Simulation of outcome data
sim_dat <- SimulateJointLongData(n.sample = n,
                                 times = times,
                                 p = 3,
                                 lambda.t = lambda.t,
                                 theta.t = theta.t,
                                 varphi.t = varphi.t,
                                 xi.t = xi.t,
                                 zeta.long = zeta.long,
                                 eta.long = eta.long,
                                 beta.long = beta.long,

                                 sd.long.trajectory = 1)

sim_data <- sim_dat$df.return %>%
  ungroup()

# How many patients die over the entire study duration?
sim_data %>%
  group_by(ID) %>%
  reframe(dies = any(Death == 1)) %>%
  pull(dies) %>%
  table()

# time of death
sim_data %>%
  group_by(ID) %>%
  dplyr::filter(Death == 1) %>%
  reframe(death_time = min(Time)) %>%
  pull(death_time) %>%
  table()

```

### Patients at risk 

```{r}
# patients at risk over time
plot_patients_at_risk(sim_data)

```

### Variance across time for longitudinal marker
The longitudinal trajectory is not-stationary, as visible from the non-constant covariance over time.  


```{r}
var_over_time <- sim_data %>% 
  group_by(Time) %>% 
  reframe(var = var(Marker))


ggplot(data = var_over_time, 
       aes(x = Time, 
           y = var)) +
  geom_line() +
  theme_bw()

# estimate_ts_covariance(sim_data, 
#                        k = 1)
# estimate_ts_covariance(sim_data, 
#                        k = 5)
# estimate_ts_covariance(sim_data, 
#                        k = 10)

```


### Expected longitudinal marker trajectory for mortal and immortal cohort 

```{r}
# expected longitudinal trajectory over time
plot_immortal_mortal_longitudinal_trajectory(sim_data)
```

### Heterogeneity in longitudinal marker trajectory 

```{r}
# patient-specific longitudinal trajectories over time
plot_heterogeneity(data = sim_data)
```

```{r}
yt <- arima.sim(list(order=c(1,0,0), ar=.5), n=500)

yt %>% length()


```


