---
title: "Semi-Competing Risks - Calibration metrics"
output: html_document
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

# {.tabset .tabset-fade .tabset-pills} 

## Setup

```{r, echo=T,message=F,warning=F,results='hide'}
library(tidyverse)
library(knitr)
library(ggpubr)

library(survival)

library(SemiCompRisks)
library(SemiCompRisksFreq)
library(splines2)

library(cmprsk)
library(survival)
library(rlist)

# for all the metrics 
library(pec)
library(pROC)
library(mccr)

library(xtable)



# use Harrison's color coding 
#color blind friendly colors from here: https://davidmathlogic.com/colorblind/#%23648FFF-%23785EF0-%23DC267F-%23FE6100-%23FFB000
cb_blue <- "#648FFF"; cb_red <- "#DC267F"; cb_purple <- "#785EF0"; cb_orange <- "#FE6100"; cb_grey <- "#CACACA"

#pink and green color-blind friendly color palette, with two colors (pink and green)
#each in two shades: light and dark
four_color_paired <- RColorBrewer::brewer.pal(n=4,name="PiYG")[c(3,4,2,1)]


#other colors used in plots
two_color_cb <- c(cb_blue,cb_red)

three_color <- c("dodgerblue","firebrick3","purple3")
three_color_cb <- c(cb_blue,cb_red,cb_purple)

#take three cb colors and reduce V from 100 to 60
three_color_cb_dark <- c("#3b5699", "#751342", "#453589")

four_color <- c("lightgray","firebrick3","purple3","dodgerblue")
four_color_cb <- c(cb_grey,cb_red,cb_purple,cb_blue)
four_color_forest <- c("dodgerblue","firebrick3","purple3","magenta")
four_color_forest_cb <- c(cb_blue,cb_red,cb_purple,cb_orange)
five_color <- c("lightgray","firebrick3","magenta","purple3","dodgerblue")
five_color_cb <- c(cb_grey,cb_red,cb_orange,cb_purple,cb_blue)

four_color_grey <- c("grey80","grey60","grey40","grey20")
five_color_grey <- c("grey80","grey60","grey40","grey20","black")

# RColorBrewer::display.brewer.all(n=4,colorblindFriendly = TRUE)
# color-blind friendly categorical colors
three_color_qual <- RColorBrewer::brewer.pal(n=3,name="Set2")
four_color_qual <- RColorBrewer::brewer.pal(n=4,name="Dark2")
five_color_qual <- RColorBrewer::brewer.pal(n=5,name="Dark2")

```


## Functions 
```{r, echo=T,message=F,warning=F,results='hide'}
h3tv_knots_temp <- c(3, 8, 15.5)
# define basis function for piecewise effect of non-terminal time on terminal time 
h3_b_fun <- function(x) splines2::bSpline(x,
                                          knots = h3tv_knots_temp,
                                          Boundary.knots = c(0,Inf),
                                          degree = 0,intercept = FALSE)

# function to plot risk profile 
plot_risk_profile <- function(tseq_pred,
                              ValPred, 
                              ValDat, 
                              i, 
                              colors = four_color_cb, 
                              t_end) {
  length_tseq <- length(tseq_pred)
  
  # format data frame 
  plot_frame <- data.frame(
    Time = rep(tseq_pred, 4),
    Probability = c(ValPred$p_neither[, i], 
                    ValPred$p_term_only[, i],
                    ValPred$p_both[, i],
                    ValPred$p_nonterm_only[, i]),
    Outcome=factor(x = c(rep("Neither",length_tseq),
                         rep("Terminal Only",length_tseq),
                         rep("Both",length_tseq),
                         rep("Nonterminal Only",length_tseq)),
                   levels = c("Neither",
                              "Terminal Only",
                              "Both",
                              "Nonterminal Only"))
  )
  
    g <- ggplot() +
    geom_area(plot_frame, 
              mapping = aes(x = Time, 
                  y = Probability, 
                  colour = Outcome, 
                  fill = Outcome)) +
    scale_color_manual(values = colors) +
    scale_fill_manual(values = colors) + 
    theme_bw() +
    labs(title = paste0("Patient ", i, collapse = ""))
    
    if(ValDat[i, ]$agvhd24 == 1) {
    trueT1 <- ValDat[i, ]$intxagvhd24_cens
    
    g <- g +
      geom_step(data = data.frame(time = c(0, trueT1, t_end), 
                                  ind = c(0, 1, 1)), 
                aes(x = time, 
                    y = ind), 
                 linetype = "dashed", 
                 color = "red")
    
  }
  if(ValDat[i, ]$dead_cens == 1) {
    trueT2 <- ValDat[i, ]$intxdeath_cens
    
     g <- g +
      geom_step(data = data.frame(time = c(0, trueT2, t_end), 
                                  ind = c(0, 1, 1)), 
                aes(x = time, 
                    y = ind), 
                linetype = "dashed", 
                 color = "black")
  }

    return(g)
}


plot_RMTL_profile <- function(tseq_pred,
                              ValPred, ValDat, 
                              i, 
                              colors = four_color_cb) {
  
  RMTL <- data.frame(time = tseq_pred, 
                     RMTL_NT = c(0, rep(NA, length(tseq_pred) - 1)), 
                     RMTL_T = c(0, rep(NA, length(tseq_pred)- 1)), 
                     RMTL_NT_T = c(0, rep(NA, length(tseq_pred) - 1)), 
                     RMTL_HA = c(0, rep(NA, length(tseq_pred) - 1))
  )
  
  for (t in tseq_pred[-1]) {
    
    tseq_temp <- ValPred$tseq[ValPred$tseq <= t]
    P_nonterm_temp <- ValPred$p_nonterm_only[seq(1, length(tseq_temp)), i]
    P_term_temp <- ValPred$p_term_only[seq(1, length(tseq_temp)), i]
    P_both_temp <- ValPred$p_both[seq(1, length(tseq_temp)), i]
    P_neither_temp <- ValPred$p_neither[seq(1, length(tseq_temp)), i]
    
    RMTL[t + 1, ]$RMTL_NT <- as.vector(c(0.5, diff(tseq_temp)[-1], 0.5) %*% 
                                         P_nonterm_temp)
    
    RMTL[t + 1, ]$RMTL_T <- as.vector(c(0.5, diff(tseq_temp)[-1], 0.5) %*% 
                                        P_term_temp) 
    
    RMTL[t + 1, ]$RMTL_NT_T <- as.vector(c(0.5, diff(tseq_temp)[-1], 0.5) %*% 
                                           P_both_temp) 
    
    RMTL[t + 1, ]$RMTL_HA <- as.vector(c(0.5, diff(tseq_temp)[-1], 0.5) %*% 
                                         P_neither_temp) 
  }
  
  RMTL_df <- RMTL %>% 
    gather(key = "name", 
           value = "RMTL", 
           -time)
  
  # plot RMTL over time 
  g <- ggplot(data = RMTL_df, 
              aes(x = time, 
                  y = RMTL,
                  color = name, 
                  fill = name)) +
    geom_area(position = "stack") +
    # geom_line() +
    theme_bw() +
    theme(legend.position = "bottom") +
    scale_color_manual(values = colors) +
    scale_fill_manual(values = colors) + 
    labs(color = "", fill = "", 
         x = "Truncation time", 
         title = paste0("Patient ", i, collapse = ""))
  
  
      if(ValDat[i, ]$agvhd24 == 1) {
    trueT1 <- ValDat[i, ]$intxagvhd24_cens
    
    g <- g +
      geom_vline(aes(xintercept = trueT1), 
                 linetype = "dashed", 
                 color = "red")
    
  }
  if(ValDat[i, ]$dead_cens == 1) {
    trueT2 <- ValDat[i, ]$intxdeath_cens
    
     g <- g +
      geom_vline(aes(xintercept = trueT2), 
                 linetype = "dashed", 
                 color = "black")
  }
    return(g)
}

```
## Data 
The data is given in terms of days. 


There are 6459 patients suffering from graft-vs-host-disease and 1823 patients die. 
816 patients die after experiencing graft-vs-host-disease.

```{r, echo=T,message=F,warning=F,results='hide'}
load("/Users/stephaniearmbruster/Desktop/Grant/R/Graft_vs_host_dataset/CIBMTR_cleaned05042923.RData")

# data check 
cibmtr %>% nrow()
cibmtr %>% colnames()

# delete missing values 
# complete case analysis for covariates of interest 
cibmtr_analysis <- cibmtr[!is.na(cibmtr$agvhd24) & !is.na(cibmtr$intxagvhd24), ]
cibmtr_analysis <- cibmtr_analysis[!is.na(cibmtr_analysis$age) & !is.na(cibmtr_analysis$stage), ]

# non-terminal event: acute graft-vs-host-disease 
cibmtr_analysis$intxagvhd24_cens %>% range()
cibmtr_analysis$agvhd24 %>% table()

# terminal event: death 
cibmtr_analysis$intxdeath_cens %>% range()
cibmtr_analysis$dead_cens %>% table()

# joint 
table(cibmtr_analysis$agvhd24, cibmtr_analysis$dead_cens)
```

## Fit shared frailty Illness-Death model 
```{r, echo=T,message=F,warning=F,results='hide'}
# define formula 
# time1 + eventindicator1 | time2 + eventindicator2
form <- Formula(intxagvhd24_cens + agvhd24 | intxdeath_cens + dead_cens ~ sexFemale+ageless20+age20to39+age40to59+stageIntermediate+stageAdvanced
                | sexFemale+ageless20+age20to39+age40to59+stageIntermediate+stageAdvanced 
                | sexFemale+ageless20+age20to39+age40to59+stageIntermediate+stageAdvanced)

cibmtr_fit_freq <- FreqID_HReg2(form, 
                               data=cibmtr_analysis,
                               model="semi-Markov", 
                               frailty=TRUE, 
                               subset=NULL)

cibmtr_fit_freq %>% summary()

# predict for a certain time window 
tseq_pred <- seq(0, 3.5, by = 0.1)
cibmtr_fit_pred <- predict(cibmtr_fit_freq, 
                              tseq = tseq_pred)
```


## Coefficients for hazard function 
```{r, echo=T,message=F,warning=F,results='show'}
HR <- round(exp(summary(cibmtr_fit_freq)$coef), 2)
HR_table <- t(apply(X = HR,
                    MARGIN = 1,
                    FUN = function(x){c(paste0(x[1],
                                               " (",x[2],
                                               ", ",x[3],")"),
                                        paste0(x[4]," (",x[5],", ",x[6],")"),
                                        paste0(x[7]," (",x[8],", ",x[9],")"))}))
colnames(HR_table) <- c("Hazard 1", 
                        "Hazard 2", 
                        "Hazard 3")

# xtable(HR_table)
```
The coefficients for the shared frailty illness death model are: 
```{r, echo=F,message=F,warning=F,results='show'}
knitr::kable(HR_table)
```




## Visualization of hazard function  
```{r, echo=T,message=F,warning=F,results='show'}
plot(cibmtr_fit_pred, plot.est = "Haz")
plot(cibmtr_fit_pred, plot.est = "CumHaz")
plot(cibmtr_fit_pred, plot.est = "Surv")

```
## Risk prediction  
No separation in training and validation data 

```{r, echo=T,message=F,warning=F,results='show'}
# extract design matric 
DesignX <- cibmtr_analysis %>% 
  dplyr::select(sexFemale, ageless20, age20to39, age40to59, stageIntermediate, stageAdvanced) %>% 
  as.matrix()

# predict individual risk profiles 
cibmtr_fit_pred <- SemiCompRisksFreq:::pred_risk_ID(tseq = tseq_pred,
                                            para = cibmtr_fit_freq$estimate,
                                            
                                            x1new = DesignX, 
                                            x2new = DesignX, 
                                            x3new = DesignX,
                                            
                                            frailty = cibmtr_fit_freq$frailty, 
                                            model = cibmtr_fit_freq$model,
                                            nP0 = cibmtr_fit_freq$nP0, 
                                            nP = cibmtr_fit_freq$nP,
                                            p3tv = 0, 
                                            h3tv_basis_func = h3_b_fun, 
                                            hazard = cibmtr_fit_freq$hazard,
                                            knots_list = cibmtr_fit_freq$knots_list,
                                            n_quad = cibmtr_fit_freq$n_quad, 
                                            quad_method = cibmtr_fit_freq$quad_method,
                                            Finv = NULL, 
                                            alpha = 0.05)
```

### Risk profiles 

```{r, echo=T,message=F,warning=F,results='show', out.width = '1000px', dpi=300}
trunc_time <- 3.5


g_list <- list()
j <- 1 

for (i in c(1, 10, 300, 500, 700, 1000)) {
  
  g <- plot_risk_profile(tseq_pred = tseq_pred, 
                         ValPred = cibmtr_fit_pred, 
                         ValDat = cibmtr_analysis, 
                         i = i, 
                         t_end = trunc_time)
  
  g_list[[j]] <- g 
  j <- j + 1
  
}

ggarrange(plotlist = g_list, 
          nrow = 2, ncol = 3, 
          common.legend = TRUE,
          legend = "bottom")
```

## ROC analysis 
```{r, echo=T,message=F,warning=F,results='show', out.width = '1000px', dpi=300}
sens_spec_list <- list()
ROC_list <- list()

j <- 1 
for (t in c(1, 2, 3)) {
  
  t_i <- which(cibmtr_fit_pred$tseq == t)
  
  valDat_temp <- cibmtr_analysis %>% 
    mutate(PP_nonterm = cibmtr_fit_pred$p_nonterm_only[t_i, ], 
           PP_term = cibmtr_fit_pred$p_term_only[t_i, ], 
           PP_both = cibmtr_fit_pred$p_both[t_i, ], 
           PP_ha = cibmtr_fit_pred$p_neither[t_i, ], 
           
           # indicator for each state 
           delta_nonterm = as.numeric(intxagvhd24_cens <= t & agvhd24 == 1 & intxdeath_cens > t), 
           delta_term = as.numeric(intxdeath_cens <= t & agvhd24 == 0 & dead_cens == 1), 
           delta_both = as.numeric(intxagvhd24_cens <= t & agvhd24 == 1 & intxdeath_cens <= t & dead_cens == 1), 
           delta_ha = as.numeric(intxagvhd24_cens > t & intxdeath_cens > t)
           )
  
  ROC_nonterm <- roc(valDat_temp$delta_nonterm, 
                    valDat_temp$PP_nonterm)
  
  ROC_term <- roc(valDat_temp$delta_term, 
                    valDat_temp$PP_term)
  
  ROC_both <- roc(valDat_temp$delta_both, 
                    valDat_temp$PP_both)
  
  ROC_neither <- roc(valDat_temp$delta_ha, 
                    valDat_temp$PP_ha)
  
  ROC_list[[j]] <- data.frame(time = t, 
                              AUC_nonterm = ROC_nonterm$auc, 
                              AUC_term = ROC_term$auc, 
                              AUC_both = ROC_both$auc, 
                              AUC_neither = ROC_neither$auc)
  
  sens_spec_list[[j]] <- data.frame(time = t, 
                              type = c(rep("Non-Terminal", length(ROC_nonterm$sensitivities)), 
                                       rep("Terminal", length(ROC_term$sensitivities)), 
                                       rep("Both", length(ROC_both$sensitivities)), 
                                       rep("Healthy and Alive", length(ROC_neither$sensitivities))
                                       ), 
                              sens = c(ROC_nonterm$sensitivities, 
                                       ROC_term$sensitivities, 
                                       ROC_both$sensitivities, 
                                       ROC_neither$sensitivities), 
                              spec = c(ROC_nonterm$specificities,
                                       ROC_term$specificities, 
                                       ROC_both$specificities,
                                       ROC_neither$specificities)
                              ) 
  
  j <- j + 1
}

ROC_df <- list.rbind(ROC_list)
sens_spec_df <- list.rbind(sens_spec_list)

g1 <- ggplot(data = sens_spec_df %>% filter(time == 1), 
       aes(y = sens, 
           x = 1 - spec, 
           color = type)) +
  geom_line() +
  facet_grid(~ type) +
  theme_bw() +
  labs(y = "Sensitivity", 
       x = "1 - Specificity", 
       color = "", 
       title = "ROC at t=1") +
  geom_abline(aes(intercept = 0, 
                  slope = 1), 
              linetype = "dashed", 
              color = "grey")

g2 <- ggplot(data = sens_spec_df %>% filter(time == 2), 
       aes(y = sens, 
           x = 1 - spec, 
           color = type)) +
  geom_line() +
  facet_grid(~ type) +
  theme_bw() +
  labs(y = "Sensitivity", 
       x = "1 - Specificity", 
       color = "", 
       title = "ROC at t=2") +
  geom_abline(aes(intercept = 0, 
                  slope = 1), 
              linetype = "dashed", 
              color = "grey")

g3 <- ggplot(data = sens_spec_df %>% filter(time == 3), 
       aes(y = sens, 
           x = 1 - spec, 
           color = type)) +
  geom_line() +
  facet_grid(~ type) +
  theme_bw() +
  labs(y = "Sensitivity", 
       x = "1 - Specificity", 
       color = "", 
       title = "ROC at t=3") +
  geom_abline(aes(intercept = 0, 
                  slope = 1), 
              linetype = "dashed", 
              color = "grey")

ggarrange(g1, g2, g3, 
          nrow = 3, 
          common.legend = TRUE, legend = "bottom")

ROC_df %>% unique()
```




## Expected calibration error  

```{r, echo=T,message=F,warning=F,results='show', out.width = '1000px', dpi=300}
ECE_list  <- list()
j <- 1

# all landmark points
for (t in tseq_pred) {
  
  t_i <- which(cibmtr_fit_pred$tseq == t)
  
  ECE_list[[j]] <- cibmtr_analysis %>% 
    mutate(PP_nonterm = cibmtr_fit_pred$p_nonterm_only[t_i, ], 
           PP_term = cibmtr_fit_pred$p_term_only[t_i, ], 
           PP_both = cibmtr_fit_pred$p_both[t_i, ], 
           PP_ha = cibmtr_fit_pred$p_neither[t_i, ], 
           
           # indicator for each state 
           delta_nonterm = as.numeric(intxagvhd24_cens <= t & agvhd24 == 1 & intxdeath_cens > t), 
           delta_term = as.numeric(intxdeath_cens <= t & agvhd24 == 0 & dead_cens == 1), 
           delta_both = as.numeric(intxagvhd24_cens <= t & agvhd24 == 1 & intxdeath_cens <= t & dead_cens == 1), 
           delta_ha = as.numeric(intxagvhd24_cens > t & intxdeath_cens > t)
           )%>% 
    summarize(
      time = t, 
      "Nonterminal" = mean(abs(PP_nonterm - delta_nonterm)), 
      "Terminal" = mean(abs(PP_term - delta_term)), 
      "Nonterminal + Terminal" = mean(abs(PP_both - delta_both)), 
      "Healthy and Alive" = mean(abs(PP_ha - delta_ha))
    )
  
  j <- j + 1 
}

ECE_df <- ECE_list %>% 
  list.rbind() %>% 
  gather(type, 
         ece, 
         -time)

# visualize over time 
ggplot(data = ECE_df, 
       aes(x = time, 
           y  = ece, 
           color = type)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  theme(legend.position = "bottom") +
  labs(x = "Time", 
       y = "Expected calibration error (ECE)", 
       color = "")

```


### Integrated expected calibration error 
```{r, echo=T,message=F,warning=F,results='show', out.width = '1000px', dpi=300}
ECE_df %>% 
  group_by(type) %>% 
  mutate(time_lag = c(0, lag(time) %>% na.omit()), 
         auc_part = ece * (time - time_lag)) %>% 
  summarize(iECE = sum(auc_part), 
            stand_iECE = iECE / 3.5)


```







